Hierarchical Time Series Forecasting in Emergency Medical Services

### 2023-12-14 journal_correspondence/response_letter/R2_response_letter_JSR.qmd

# Associate Editor

We received two very good reviews that came up with justified concern and mixed recommendations. Let me add additional comments.

## AE comments

_Comment 1. The authors need to score higher on the economic insights. Their example did not impress me, and I was hoping to receive more intuition as to why their approach works better._

  **Response:** Thank you for your feedback. In this paper, we introduce a comprehensive framework for probabilistic forecast reconciliation in Emergency Medical Services. Our work not only outlines the rationale behind the proposed approach's suitability for this specific context but also rigorously evaluates its effectiveness through statistical forecast accuracy metrics, as presented in Table 3 and Figure 4. In response to the comment on insights into the economic implications of our approach, we have added the following paragraphs to the paper to further discuss this point:

\textcolor{teal}{Our research establishes a strong basis for future investigations and practical implementation in EMS. Leveraging the hierarchical and grouped structure of demand time series, EMS can use this advanced forecasting framework to generate coherent point and probabilistic forecasts, making the most of all available data at every level of the hierarchy. We acknowledge that a forecast serves a greater purpose beyond its mere existence, ideally enabling the best utility in terms of efficient allocation of medical services, response time, and cost, all informed by the forecast. While we fully appreciate the importance of evaluating forecast quality based on its impact on decision-making processes, it is essential to address the data requirements and methodology involved in measuring this impact. For a comprehensive assessment of the forecasts' implications, access to additional data beyond ambulance demand, covering various decision types, capacity information, constraints in the decision system, and more, becomes necessary. This additional data would offer valuable insights into the specific decisions relying on the forecasts, resulting in a more accurate evaluation of their impact on medical services. Furthermore, measuring the actual impact of forecasts would necessitate an approach that goes beyond forecasting itself. This would involve developing and implementing simulation models capable of replicating decision-making processes based on the forecast inputs. These simulation models would then evaluate the quality of the final decisions, taking into consideration the utilities that are particularly significant in the context of EMS.}

_Comment 2. The authors miss opportunities. I appreciate that the authors summarize previous research in Table 1. Yet, there is nothing in Table 1 that helps the reader understand why the current research has the limitations that the authors state on page 8._

  **Response:** Thank you for your comment. In Table 1, we present three dimensions that highlights the limitations identified in prior research. These dimensions include reconciliation, probabilistic forecasting, and reproducibility. It is noteworthy that there is a notable gap in the existing literature regarding the consideration of probabilistic forecast reconciliation while adhering to the principles of reproducibility. The first column in the table relates to our research, indicating 'YES' in all these dimensions, whereas previous studies have been characterized by a 'NO' in this regard. Table 1 in the previous version had only Reconciliation, and Probabilistic. We now added the reproducibility column to the table, aligning with limitations discussed in the literature section.

_Comment 3. I highly appreciate that the authors intend to provide data and code via their GitHub repository. No doubt, the resulting transparency is a major strength of this manuscript. Yet, the authors could still do a much better job in outlining to the readers that the authors have done a good job._

  **Response:** Thank you. We would like to emphasize that our commitment to transparency and reproducibility goes beyond merely sharing data and code. Our entire paper is authored in R using [Quarto](https://quarto.org/), setting it apart as a uniquely transparent and reproducible workflow and this could be used as an outstanding example of a transparent and reproducible research in academic communities. In this way, we provide not only the code and data but also the entire narrative of the research process. To further highlight that, we have expanded the the reproducibility section in the revised manuscript:

\textcolor{teal}{To enhance transparency and reproducibility, we not only provide data and the code, but also the entire paper that is written in R using [Quarto](https://quarto.org/). All materials to reproduce this paper is available at [github.com/bahmanrostamitabar/forecasting-emergency-medicine](https://github.com/bahmanrostamitabar/forecasting-emergency-medicine). The repository contains the raw data, all R scripts used in experiments, the results used in the paper, as well as the quarto files for producing this paper. Full instructions are provided in the repository.}

_Comment 4.I felt that the authors did not address the following concern that I outlined in the last round of review: “It would also help to have a rather detailed example, maybe even a numerical example, to better describe why separate forecasts at different levels can cause problems. Right now, it sounds plausible that such problems occur, but you are still rather vague concerning the details of these problems.”_

  **Response:** Thank you. We have now added a more detailed but simple example in the introduction based on the ambulance data and added the following text to the revised version of the paper:

\textcolor{teal}{To illustrate the problem, let's consider a very simple example where we have an EMS provider with a national level of governance, and two regions (A and B), each with a health board and a station. There is a total national budget to be split between the regions in proportion to the forecast number of incidents in each region. The two regions have very different incident patterns, and so must be forecast using different models. However, the data are noisy at regional level, so the national forecasts are best obtained by summing the demand from the two regions. The resulting national forecasts are not equal to the sum of the regional forecasts, and so are not coherent. In fact, the national forecasts show a decreasing trend in demand, and so the national governing body decides to cut the budget for the next year. But neither of the regional forecasts shows a trend, and so the regions argue that the budget cut is unfair. In addition, Region A has much more variable demand than Region B, and so to cope with periods of peak demand, Region A needs to hold more resources in reserve. So the budget distribution needs to be made in a way that ensures the probability of each region being unable to meet demand is equal. Our solution to this problem is to use a hierarchical forecasting approach that ensures the forecasts are probabilistically coherent. Then any trends or other forecast characteristics at national level will also be reflected in the regional forecasts, and the probabilistic forecasts allow for the different levels of uncertainty in the two regions. Budget can be allocated by controlling the probability of demand exceeding available resources, rather than being simply in proportion to the expected demand.}

_Comment 5. I also had the following comments in the previous round of review: “Make sure that all tables and figures are self-contained” and “If possible, display the number of observations in all tables and figures with statistical results”. Unfortunately, Figure 1 is not self-contained (e.g., abbreviations are not explained). Figure 2 could also contain the correlation and the number of observations._

  **Response:** Thank you for the comment. We have anonymized the name of the health board using two letters (e.g.) that hold no meaningful association, so these are not abbreviations. We have clarified this in the caption of Figure 1.

  With regard to Figure 2, we appreciate the interest in examining correlations. However, we would like to clarify that the purpose of the scatter plot in Figure 2 is not to show the correlation. Each point on the plot corresponds to an individual time series, and the primary objective of this plot is to facilitate a clear understanding of time series features in our dataset such as the strength of trend and seasonality. Introducing correlation values in this specific plot may not add meaningful insights due to the nature of the data representation.

_Comment 6. I was also hoping for an example that not only describes the problem that the authors solve, but that allows us to better understand the basic idea behind the authors’ solution._

  **Response:** Thank you. We've addressed this comment by including a detailed example in the introduction. This addition is designed to clarify the problem and elaborate on the idea behind the forecast reconciliation. We added the following paragraph to the revised version of the paper:

\textcolor{teal}{To illustrate the problem, let's consider a very simple example where we have an EMS provider with a national level of governance, and two regions (A and B), each with a health board and a station. There is a total national budget to be split between the regions in proportion to the forecast number of incidents in each region. The two regions have very different incident patterns, and so must be forecast using different models. However, the data are noisy at regional level, so the national forecasts are best obtained by summing the demand from the two regions. The resulting national forecasts are not equal to the sum of the regional forecasts, and so are not coherent. In fact, the national forecasts show a decreasing trend in demand, and so the national governing body decides to cut the budget for the next year. But neither of the regional forecasts shows a trend, and so the regions argue that the budget cut is unfair. In addition, Region A has much more variable demand than Region B, and so to cope with periods of peak demand, Region A needs to hold more resources in reserve. So the budget distribution needs to be made in a way that ensures the probability of each region being unable to meet demand is equal. Our solution to this problem is to use a hierarchical forecasting approach that ensures the forecasts are probabilistically coherent. Then any trends or other forecast characteristics at national level will also be reflected in the regional forecasts, and the probabilistic forecasts allow for the different levels of uncertainty in the two regions. Budget can be allocated by controlling the probability of demand exceeding available resources, rather than being simply in proportion to the expected demand.}

### 2023-12-14 manuscript/01-introduction.qmd
finalised comments of AE

Demand forecasts at EMS are typically required at multiple levels of an organization to inform various planning and decision-making processes [@hulshof2012taxonomic]. There are some planning processes at the national level (strategic and long-term) such as workforce resource planning and budgeting; sub-national, regional, or healthcare level (tactical and medium-term) such as temporary capacity expansions, resource sharing; and hospital or station level (operational and short-term) such as planning rosters for staff and ambulance deployment. Demand forecasts might also be required at different levels for a specific area of interest such as the nature of demand or the priority level. Moreover, the time series data in EMS has an inherent hierarchical and grouped structure to support such forecasting requirements. Demand for emergency medical services at the national level can be disaggregated in a geographical hierarchy into sub-national, regions, health boards, and stations/hospitals, or divided into groups such as the nature of incidents or demand priority. \textcolor{teal}{Forecasts produced at both higher and lower levels of hierarchies are necessary for effective decision-making in EMS. For example, control area EMS forecasts can inform strategic decisions about how to allocate limited resources to lower levels, such as health boards and stations/hospitals. At the lower levels, hospitals or ambulance stations could use such forecasts to plan for staffing and resource allocation, ambulance dispatching, staff-to-shift assignment, staff rescheduling based on the anticipated volume and priority and nature of incidents. Additionally, generating forecasts at lower levels could potentially improve the accuracy of the high-level forecasts, by providing more detailed information on the nature and priority of incidents. This could help to identify patterns in demand that may not be apparent at the higher level. Therefore, employing forecasting techniques that consider the hierarchical and/or grouped patterns of time series in EMS aligns naturally, offering the possibility to enhance forecast accuracy and facilitate coordination.}

To illustrate the problem, let's consider a very simple example where we have an EMS provider with a national level of governance, and two regions (A and B), each with a health board and a station. There is a total national budget to be split between the regions in proportion to the forecast number of incidents in each region. The two regions have very different incident patterns, and so must be forecast using different models. However, the data are noisy at regional level, so the national forecasts are best obtained by summing the demand from the two regions. The resulting national forecasts are not equal to the sum of the regional forecasts, and so are not coherent. In fact, the national forecasts show a decreasing trend in demand, and so the national governing body decides to cut the budget for the next year. But neither of the regional forecasts shows a trend, and so the regions argue that the budget cut is unfair. In addition, Region A has much more variable demand than Region B, and so to cope with periods of peak demand, Region A needs to hold more resources in reserve. So the budget distribution needs to be made in a way that ensures the probability of each region being unable to meet demand is equal. Our solution to this problem is to use a hierarchical forecasting approach that ensures the forecasts are probabilistically coherent. Then any trends or other forecast characteristics at national level will also be reflected in the regional forecasts, and the probabilistic forecasts allow for the different levels of uncertainty in the two regions. Budget can be allocated by controlling the probability of demand exceeding available resources, rather than being simply in proportion to the expected demand.

Despite a large number of studies dedicated to forecasting for EMS [@mingliterature2022; @gul2020exhaustive; @ibrahim2016modeling; @wargon2009systematic], the hierarchical data structure has been largely ignored, and the main focus has been on producing independent (base) forecasts at a single level. Generating independent forecasts can result in a lack of consistency and coordination, and therefore leads to less effective planning and decision making. \textcolor{teal}{For example, suppose the annual budget for the whole organization is to be allocated to each area health board based on demand within the corresponding area. If the forecasts are incoherent, there is a mismatch between the total resources to be allocated, and the resources allocated to each area. Further, suppose the budget also needs to take account of the nature of incidents that occur within each area, with more money allocated for some types of incidents than others. Then we need forecasts of the demand disaggregated by nature of incidents and health board, but these data are often too noisy to forecast directly. Forecast reconciliation addresses these problems by ensuring the forecasts are coherent across all aggregated and disaggregated series (avoiding mismatches), and by using the signal in the aggregated data to allow forecasting of highly disaggregated data (allowing better targetting of the budget).}

>>>>

Demand forecasts at EMS are typically required at multiple levels of an organization to inform various planning and decision-making processes [@hulshof2012taxonomic]. There are some planning processes at the national level (strategic and long-term) such as workforce resource planning and budgeting; sub-national, regional, or healthcare level (tactical and medium-term) such as temporary capacity expansions, resource sharing; and hospital or station level (operational and short-term) such as planning rosters for staff and ambulance deployment. Demand forecasts might also be required at different levels for a specific area of interest such as the nature of demand or the priority level. Moreover, the time series data in EMS has an inherent hierarchical and grouped structure to support such forecasting requirements. Demand for emergency medical services at the national level can be disaggregated in a geographical hierarchy into sub-national, regions, health boards, and stations/hospitals, or divided into groups such as the nature of incidents or demand priority. Forecasts produced at both higher and lower levels of hierarchies are necessary for effective decision-making in EMS. For example, control area EMS forecasts can inform strategic decisions about how to allocate limited resources to lower levels, such as health boards and stations/hospitals. At the lower levels, hospitals or ambulance stations could use such forecasts to plan for staffing and resource allocation, ambulance dispatching, staff-to-shift assignment, staff rescheduling based on the anticipated volume and priority and nature of incidents. Additionally, generating forecasts at lower levels could potentially improve the accuracy of the high-level forecasts, by providing more detailed information on the nature and priority of incidents. This could help to identify patterns in demand that may not be apparent at the higher level. Therefore, employing forecasting techniques that consider the hierarchical and/or grouped patterns of time series in EMS aligns naturally, offering the possibility to enhance forecast accuracy and facilitate coordination.

\textcolor{teal}{To illustrate the problem, let's consider a very simple example where we have an EMS provider with a national level of governance, and two regions (A and B), each with a health board and a station. There is a total national budget to be split between the regions in proportion to the forecast number of incidents in each region. The two regions have very different incident patterns, and so must be forecast using different models. However, the data are noisy at regional level, so the national forecasts are best obtained by summing the demand from the two regions. The resulting national forecasts are not equal to the sum of the regional forecasts, and so are not coherent. In fact, the national forecasts show a decreasing trend in demand, and so the national governing body decides to cut the budget for the next year. But neither of the regional forecasts shows a trend, and so the regions argue that the budget cut is unfair. In addition, Region A has much more variable demand than Region B, and so to cope with periods of peak demand, Region A needs to hold more resources in reserve. So the budget distribution needs to be made in a way that ensures the probability of each region being unable to meet demand is equal. Our solution to this problem is to use a hierarchical forecasting approach that ensures the forecasts are probabilistically coherent. Then any trends or other forecast characteristics at national level will also be reflected in the regional forecasts, and the probabilistic forecasts allow for the different levels of uncertainty in the two regions. Budget can be allocated by controlling the probability of demand exceeding available resources, rather than being simply in proportion to the expected demand.}

Despite a large number of studies dedicated to forecasting for EMS [@mingliterature2022; @gul2020exhaustive; @ibrahim2016modeling; @wargon2009systematic], the hierarchical data structure has been largely ignored, and the main focus has been on producing independent (base) forecasts at a single level. Generating independent forecasts can result in a lack of consistency and coordination, and therefore leads to less effective planning and decision making.

### 2023-12-14 manuscript/main.qmd
finalised comments of AE

R code to produce all results in this paper is available at [github.com/bahmanrostamitabar/forecasting-emergency-medicine](https://github.com/bahmanrostamitabar/forecasting-emergency-medicine). The repository contains the raw data, all R scripts used in experiments as well as the results used in this paper, along with the raw files for producing this paper. Full instructions are provided in the repository.

>>>>

To enhance transparency and reproducibility, we not only provide data and the code, but also the entire paper that is written in R using [Quarto](https://quarto.org/). All materials to reproduce this paper is available at [github.com/bahmanrostamitabar/forecasting-emergency-medicine](https://github.com/bahmanrostamitabar/forecasting-emergency-medicine). The repository contains the raw data, all R scripts used in experiments, the results used in the paper, as well as the quarto files for producing this paper. Full instructions are provided in the repository.
